---
title: "Introduction to scToppR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package functions as an API wrapper to [ToppGene](https://toppgene.cchmc.org/). It takes a file from Seurat's FindAllMarkers, Presto's Wilcoxauc functions, or similarly formatted data that contains columns of genes, groups of cells (clusters or celltypes), avg log fold changes, and p-values. 

As an introduction, this vignette will work with the FindAllMarkers output from Seurat's PBMC 3k clustering tutorial: [https://satijalab.org/seurat/articles/pbmc3k_tutorial.html](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

You can follow that tutorial and get the markers file from this line:

```
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)
```
Alternatively, this markers table is included in the scToppR package:
```{r setup}
#library(scToppR)
devtools::load_all()
data("pbmc.markers")
head(pbmc.markers)
```


With this data we can run the function `toppFun` to get results from ToppGene.

```{r toppFun_1}
toppData <- toppFun(markers = pbmc.markers,
                    topp_categories = NULL, 
                    cluster_col = "cluster", 
                    gene_col = "gene")

```

Here it is important to tell toppFun the names of the relevant columns for clusters and genes. Additionally, you can run toppFun on all ToppGene categories by setting topp_categories to NULL. You may also provide 1 or more specific categories as a list. To see all ToppGene categories, you can also use the function get_ToppCats():

```{r topp_cats}
get_ToppCats()
```

You can also set additional parameters in the toppFun function, please check the documentation for more information.

The results of toppFun are organized into a data frame as such:

```{r toppData_out}
head(toppData)

```

## Plotting

scToppR can automatically create DotPlots for each ToppGene category. Simply run:

```{r toppPlot_1}
plots <- toppPlot(toppData, category = "GeneOntologyMolecularFunction", clusters = NULL)
plots[1]
```

This will create a list of plots for all categories and clusters requested. Here, the category "GenoOntologyMolecularFunction" was requested, and the `clusters` parameter was left NULL as default. If either `category` or `clusters` are NULL, then all available options are used. For example, the output here creates a list of plots for each cluster for the "GenoOntologyMolecularFunction". If multiple categories and all clusters are used, it will create a list of lists. This can get unwieldly, so the function can automatically save each plot as well, using the `save = TRUE` parameter. This will automatically save each file in the format: `{category}_{cluster}_dotplot.pdf`

scToppR can also create balloon plots showing overlapping terms between all clusters.

```{r toppBalloon}
toppBalloon(toppData, categories = "GeneOntologyMolecularFunction")
```

This function also has a save parameter, which will automatically save plots, which is helpful if multiple categories are visualized.


## Saving 

scToppR will also automatically save the results of the ToppGene query. By default it will save separate files for each cluster. To save as one large file, set the parameter `split = FALSE`. It will also save all files as Excel spreadsheets, but this can be changed using the `format` parameter--it must be one of `c("xlsx", "csv", "tsv")`.

```{r}
toppSave(toppData, filename = "PBMC", split = TRUE, format = "xslx")
```

