---
title: "Introduction to scToppR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package functions as an API wrapper to [ToppGene](https://toppgene.cchmc.org/). It takes a file from Seurat's FindAllMarkers, Presto's Wilcoxauc functions, or similarly formatted data that contains columns of genes, groups of cells (clusters or celltypes), avg log fold changes, and p-values. 

As an introduction, this vignette will work with the FindAllMarkers output from Seurat's PBMC 3k clustering tutorial: [https://satijalab.org/seurat/articles/pbmc3k_tutorial.html](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

You can follow that tutorial and get the markers file from this line:

```
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)
```
Alternatively, this markers table is included in the scToppR package:
```{r setup}
#library(scToppR)
devtools::load_all()
data("pbmc.markers")
head(pbmc.markers)
```


With this data we can run the function `toppFun` to get results from ToppGene.

```{r toppFun_1}
toppData <- toppFun(markers = pbmc.markers,
                    topp_categories = NULL, 
                    cluster_col = "cluster", 
                    gene_col = "gene")

```

Here it is important to tell toppFun the names of the relevant columns for clusters and genes. Additionally, you can run toppFun on all ToppGene categories by setting topp_categories to NULL. You may also provide 1 or more specific categories as a list. To see all ToppGene categories, you can also use the function get_ToppCats():

```{r topp_cats}
get_ToppCats()
```

You can also set additional parameters in the toppFun function, please check the documentation for more information.

The results of toppFun are organized into a data frame as such:

```{r toppData_out}
head(toppData)

```

